---
title: "Yojoa Remote Sensing Methods"
author: "B Steele, Matt Ross"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(ggthemes)
library(readr)

is_dir <- "data/in-situ/"
rs_dir <- "data/upstreamRS/"
match_dir <- "data/matchups/"
models_dir <- "data/models/"
```

# Methods

```{r load data, include = F}
stack = read_csv(file.path(rs_dir, 'yojoa_corr_rrs_met_v2023-06-15.csv')) %>% 
  filter(date < ymd('2023-01-01'))
secchi = read_csv(file.path(is_dir, 'Secchi_completedataset.csv')) %>% 
  mutate(secci = as.numeric(secchi)) %>% 
  filter(!is.na(secchi)) %>% 
  mutate(location = gsub(' ', '', location))
match5 = read_csv(file.path(match_dir, 'fiveDay_LS-Secchi_matchups_n237.csv'))
match71 = read_csv(file.path(match_dir, 'multiDay_LS-Secchi_matchups_n238.csv'))
```

## Landsat Stack

Median Landsat Collection 2 Surface Reflectance (Masek, et al., 2006, Vermote, et al., 2016) values (Rrs) were obtained for the 18 sampling locations in Lake Yojoa following the methods described in Topp, et al. (2021). The Landsat stack was comprised of `r length(unique(stack$system.index))` unique scenes comprised of data from Landsat 5, 7, 8, and 9 (Table 1). This stack contained `r nrow(stack)` Rrs summaries (Table 2) at the 18 locations between the dates of `r min(stack$date)` and `r max(stack$date)`. Minor adaptations were made for the transition for Collection 1 to Collection 2 Landsat data to account for differences in scaling factors between collections. Rrs summaries included only 'confident' water pixels as defined by the dynamic surface water extent algorighm (Jones, 2019). Data were filtered for reasonable values for water reflectance (-0.01 \< Rrs \< 0.2) for all bands (Blue, Red, Green, Near Infrared, Shortwave Infrared 1, Shortwave Infrared 2). Inter-mission handoff coefficients to standardize Rrs values due to slight changes in sensors and atmospheric correction between missions (Gardner, et al. 2021) were calculated based on Landsat Collection 2 data acquired from all lakes greater than 25 hectacres within Guatemala, Honduras, and El Salvador (described in *Regional Handoff Coefficients* below). All Landsat data were acquired using the Google Earth Engine Python Application Programming Interface (API) in RStudio version 2023.03.0, R version 4.2.3 (R Core Team 2023), and Python version 3.8 (Python Software Foundation, <https://www.python.org/>).

Landsat data acquisition was completed in the scripts within the 'landsat_c2' folder. Within this folder there are two subfolders of scripts: `literateCode` which contains the scripts used to acquire the Yojoa stack and `forHandoffs` which contains the scripts used to acquire the regional remote sensing stack for the handoff calculations. Landsat data collation was completed in the script `programs/1_RS_collate_harmonize.Rmd`. Regional remote sensing data were collated in the script `programs/3_regionalRS_collate_harmonize.Rmd`. Handoff calculations were made in the script `programs/4_regional_handoff_calcs.Rmd` and were applied in the script `programs/5_apply_handoff_coefficients.Rmd`.

```{r getCount, echo = F}
# save function to get row count
getCount <- function(dataset, miss, countType) {
  if (miss == "all") {
    df <- dataset
  } else {
    df <- dataset %>%
      filter(mission == miss)
  }
  if (countType == "imageDate") {
    length(unique(df$system.index))
  } else {
    if (countType == "matches") {
      nrow(df)
    } else {
      print("oops")
    }
  }
}
```

| Landsat Mission | Number of Unique Landsat Scenes               | Number of Unique Landsat Scenes with ±5 day matchups | **Number of Unique Landsat Scenes with ±7/1 day matchups** |
|------------------|------------------|------------------|------------------|
| Landsat 5       | `r getCount(stack, 'LANDSAT_5', 'imageDate')` | `r getCount(match5, 'LANDSAT_5', 'imageDate')`       | `r getCount(match71, 'LANDSAT_5', 'imageDate')`            |
| Landsat 7       | `r getCount(stack, 'LANDSAT_7', 'imageDate')` | `r getCount(match5, 'LANDSAT_7', 'imageDate')`       | `r getCount(match71, 'LANDSAT_7', 'imageDate')`            |
| Landsat 8       | `r getCount(stack, 'LANDSAT_8', 'imageDate')` | `r getCount(match5, 'LANDSAT_8', 'imageDate')`       | `r getCount(match71, 'LANDSAT_8', 'imageDate')`            |
| Landsat 9       | `r getCount(stack, 'LANDSAT_9', 'imageDate')` | `r getCount(match5, 'LANDSAT_9', 'imageDate')`       | `r getCount(match71, 'LANDSAT_9', 'imageDate')`            |
| Total           | `r getCount(stack, 'all', 'imageDate')`       | `r getCount(match5, 'all', 'imageDate')`             | `r getCount(match71, 'all', 'imageDate')`                  |

: Table 1. Summary of number of unique Landsat scenes in the full Landsat stack, the ±5 day matchup dataset, and the ±7/1 day dataset.

| Landsat Mission | Number of Valid Location-Landsat Images     | Number of Valid Location-Landsat Images in ±5 day matchups | **Number of** Valid Location-Landsat Images in **±7/1 day matchups** |
|------------------|------------------|------------------|------------------|
| Landsat 5       | `r getCount(stack, 'LANDSAT_5', 'matches')` | `r getCount(match5, 'LANDSAT_5', 'matches')`               | `r getCount(match71, 'LANDSAT_5', 'matches')`                        |
| Landsat 7       | `r getCount(stack, 'LANDSAT_7', 'matches')` | `r getCount(match5, 'LANDSAT_7', 'matches')`               | `r getCount(match71, 'LANDSAT_7', 'matches')`                        |
| Landsat 8       | `r getCount(stack, 'LANDSAT_8', 'matches')` | `r getCount(match5, 'LANDSAT_8', 'matches')`               | `r getCount(match71, 'LANDSAT_8', 'matches')`                        |
| Landsat 9       | `r getCount(stack, 'LANDSAT_9', 'matches')` | `r getCount(match5, 'LANDSAT_9', 'matches')`               | `r getCount(match71, 'LANDSAT_9', 'matches')`                        |
| Total           | `r getCount(stack, 'all', 'matches')`       | `r getCount(match5, 'all', 'matches')`                     | `r getCount(match71, 'all', 'matches')`                              |

: Table 2. Summary of number of valid Landsat observations for individual locations in Yojoa in the full Landsat stack, the ±5 day matchup dataset, and the ±7/1 day dataset.

```{r, echo = F}
getSiteCount <- function(dataset, site) {
  if (site == "all") {
    df <- dataset
    nrow(df)
  } else {
    df <- dataset %>%
      filter(location == site)
    nrow(df)
  }
}
```

| Site  | Number of Secchi Measurements   | Number of Secchi-Landsat Matchups in ±5 day dataset | Number of Secchi-Landsat Matchups in ±7/1 day dataset. |
|------------------|------------------|------------------|------------------|
| A     | `r getSiteCount(secchi, 'A')`   | `r getSiteCount(match5, 'A')`                       | `r getSiteCount(match71, 'A')`                         |
| B     | `r getSiteCount(secchi, 'B')`   | `r getSiteCount(match5, 'B')`                       | `r getSiteCount(match71, 'B')`                         |
| C     | `r getSiteCount(secchi, 'C')`   | `r getSiteCount(match5, 'C')`                       | `r getSiteCount(match71, 'C')`                         |
| D     | `r getSiteCount(secchi, 'D')`   | `r getSiteCount(match5, 'D')`                       | `r getSiteCount(match71, 'D')`                         |
| E     | `r getSiteCount(secchi, 'E')`   | `r getSiteCount(match5, 'E')`                       | `r getSiteCount(match71, 'E')`                         |
| F     | `r getSiteCount(secchi, 'F')`   | `r getSiteCount(match5, 'F')`                       | `r getSiteCount(match71, 'F')`                         |
| G     | `r getSiteCount(secchi, 'G')`   | `r getSiteCount(match5, 'G')`                       | `r getSiteCount(match71, 'G')`                         |
| H     | `r getSiteCount(secchi, 'H')`   | `r getSiteCount(match5, 'H')`                       | `r getSiteCount(match71, 'H')`                         |
| I     | `r getSiteCount(secchi, 'I')`   | `r getSiteCount(match5, 'I')`                       | `r getSiteCount(match71, 'I')`                         |
| J     | `r getSiteCount(secchi, 'J')`   | `r getSiteCount(match5, 'J')`                       | `r getSiteCount(match71, 'J')`                         |
| K     | `r getSiteCount(secchi, 'K')`   | `r getSiteCount(match5, 'K')`                       | `r getSiteCount(match71, 'K')`                         |
| L     | `r getSiteCount(secchi, 'L')`   | `r getSiteCount(match5, 'L')`                       | `r getSiteCount(match71, 'L')`                         |
| M     | `r getSiteCount(secchi, 'M')`   | `r getSiteCount(match5, 'M')`                       | `r getSiteCount(match71, 'M')`                         |
| N     | `r getSiteCount(secchi, 'N')`   | `r getSiteCount(match5, 'N')`                       | `r getSiteCount(match71, 'N')`                         |
| O     | `r getSiteCount(secchi, 'O')`   | `r getSiteCount(match5, 'O')`                       | `r getSiteCount(match71, 'O')`                         |
| P     | `r getSiteCount(secchi, 'P')`   | `r getSiteCount(match5, 'P')`                       | `r getSiteCount(match71, 'P')`                         |
| Q     | `r getSiteCount(secchi, 'Q')`   | `r getSiteCount(match5, 'Q')`                       | `r getSiteCount(match71, 'Q')`                         |
| R     | `r getSiteCount(secchi, 'R')`   | `r getSiteCount(match5, 'R')`                       | `r getSiteCount(match71, 'R')`                         |
| Total | `r getSiteCount(secchi, 'all')` | `r getSiteCount(match5, 'all')`                     | `r getSiteCount(match71, 'all')`                       |

: Table 3. Secchi measurement counts per site in the *in situ* record, the ±5-day matchup dataset,and the ±7/1-day matchup dataset.

#### Regional Handoff Coefficients

Median Landsat Collection 2 Surface Reflectance values were obtained for the Chebyshev center (Yang, 2020) for all lakes greater than 25 hectares within or touching the boundaries of the countries of Guatemala, Honduras, and El Salvador from the spatial dataset produced by Messager, et al. (2016). Acquisition of these data followed the same methods as those for Lake Yojoa described above.

Centers of the lakes from the spatial dataset were calculated in the GEE IDE using the method from Yang (2020) with the script `GEE_scripts/guat_elsal_hon_getCenters.js`.

## Secchi-RS matchups

Secchi data were matched with records from the Landsat remote sensing stack at a number of time windows: same, one, two, three, and five day windows. In addition to these standard windows, we employed a variable window defined by local knowledge. In this method, we allowed for matchups up to ±7 days for all months where the conditions of the lake are relatively consistent and in October and November, months that often have sudden clarity changes, only matches ±1 day were permitted. Windows up to ±7 days have yielded reasonable results when paired with remote sensing data (Kloiber, et al. 2002). We avoided discrete over-matching of our data by assuring that each satellite overpass instance was only paired with the nearest-in-time Secchi measurement and that each discrete Secchi measurement was only paired with a single, nearest-in-time valid satellite overpass. Only matchup datasets of ±5 days and ±7/1 produced reasonable models and are the only datasets summarized in this methods overview.

Matchups were completed in the script `programs/7_LSC2_secchi_matchups.Rmd`.

## Climate Data

Climate data were aqcuired from the ERA5 dataset (Muñoz, 2019) in the Google Earth Engine Code Editor (Gorelick, 2017) from a single data point at the approximate geographical center of Lake Yojoa (14.8768°N, 87.9791°W) for all available data. ERA5 data provide daily modeled values for an extensive list of parameters; however, we only used total precipitation, mean air temperature, total solar radiation, and mean wind speed in our analysis. These data were aggregated for the previous 3, 5, and 7 days.

Climate data were accessed and summarized in the GEE IDE using the script `GEE_scripts/era5_download.js`. ERA5 data were collated and aggregated in the script `programs/2_Process_Summarize_ERA5.Rmd`. Data were paired with the Landsat stack in the script `programs/6_add_era5_data.Rmd`.

## xgboost Model

We used the R package {xgboost} (Chen, et al. 2023) to develop the best performing gradient tree boost algorithm for these data. Model features were median Rrs values for the blue, green, red and near infrared bands, the ratio of red to green, blue to green, red to blue, green to red, total solar radiation, maximum air temperature, mean air temperature, minimum air temperature, total precipitation, and mean wind speed. Various input feature window combinations were tested, including providing the program with the previous day's meteorology as well as a summary of the previous 5 or 7 days.

We experimented with two data handling methods with varying stringency. Both methods used train-test-validation datasets, where the train and test were provided for model development and the validation were hold out data to test performance independently of model development. The first method split the data as 60% of matchups as the training dataset, 20% for testing, and 20% for validation (`programs/8_xgboost_stringent.Rmd`). The very stringent method split the Landsat image-dates with the same proportions, so that there was no image-date crossover between the train, test, or validation dataset (`programs/9_xgboost_very_stringent.Rmd`).

Because we were particularly interested in performance at the upper end of Secchi (higher clarity), we also tried an xgboost model that weighted higher Secchi matchups as more important (`programs/10_xgboost_higher_secchi.Rmd`). This script uses the same data handling stringency as the 'very stringent\` method.

To select the best optimal xgboost hyperparameters, we used a grid search method partitioning the top 20 performing models as measured by lowest RMSE. From these models, we selected the booster that had the lowest RMSE and a train-test RMSE that was within 0.15m to avoid selecting an overfit model. If no models met these conditions the one with the closest train-test RMSE was selected as the optimal {xgboost} model.

## Stepwise Regression Model

We used the R packages {caret} (Kuhn, 2008) and {leaps} (Lumley, 2020) to perform backwards stepwise regression. We tested the same combinations of matchup windows and meteorological data as with the {xgboost} package, except all input data were normalized to values between 0 and 1 using a min-max scaling method. Because this method uses cross-wise validation (10 times cv), no test data were provided to the model. 70% of the data were used in model development and the remaining 30% were holdout data to examine the results independently. Data partitioning was completed by total number of matchups not by image date.

## Time series of estimated Secchi

The time series of Secchi estimates from the four best performing models (one from each of the above models) have been calculated in a single file (`data/landsat_estimations/Yojoa_LS-derived_Secchi_estimates.csv`, created in `programs/12_apply_models.Rmd`). The four column names indicate the model used to create the Secchi estimates. Keep in mind that input values of the features (the band and met data that define the model) that are outside of the range of values used to train the model can create equally out-of-range Secchi estimations in the stepwise regression model. In general, {xgboost} handles these outliers better, but a very conservative and cautious approach dictates that Secchi estimations made with input values that are beyond the range of those in model development should be treated with skepticism.

------------------------------------------------------------------------

# Acknowledgements

Landsat Collection 2 Level 2 Science Products courtesy of the U.S. Geological Survey.

------------------------------------------------------------------------

# Citations

Chen T, He T, Benesty M, Khotilovich V, Tang Y, Cho H, Chen K, Mitchell R, Cano I, Zhou T, Li M, Xie J, Lin M, Geng Y, Li Y, Yuan J (2023). \_xgboost: Extreme Gradient Boosting\_. R package version 1.7.3.1, \<[[https://CRAN.R-project.org/package=xgboost\\\\](https://CRAN.R-project.org/package=xgboost){.uri}](%5Bhttps://CRAN.R-project.org/package=xgboost%5D(https://CRAN.R-project.org/package=xgboost)%7B.uri%7D){.uri}\>.

Gardner, J. R., Yang, X., Topp, S. N., Ross, M. R.. V., Altenau, E. H., & Pavelsky, T. M. (2021). The color of rivers. *Geophysical Research Letters*, 48, e2020GL088946. <https://doi.org/10.1029/2020GL088946>

Gorelick, N., Hancher, M., Dixon, M., Ilyushchenko, S., Thau, D., & Moore, R. (2017). Google Earth Engine: Planetary-scale geospatial analysis for everyone. *Remote sensing of Environment*, *202*, 18-27.

Jones, J. (2019). Improved automated detection of subpixel-scale inundation-revised dynamic surface water extent (DSWE) Partial surface water tests. Remote Sensing, **11**(4), 374. <https://doi.org/10.3390/rs11040374>

Kuhn, M. (2008). Building Predictive Models in R Using the caret Package. Journal of Statistical Software, 28(5), 1--26. <https://doi.org/10.18637/jss.v028.i05>

Kloiber, S. M., Brezonik, P. L., Olmanson, L. G., & Bauer, M. E. (2002). A procedure for regional lake water clarity assessment using Landsat multispectral data. *Remote sensing of Environment*, *82*(1), 38-47.

Lumley, T. (2020). \_leaps: Regression Subset Selection\_. R package version 3.1, <https://CRAN.R-project.org/package=leaps>

Muñoz Sabater, J., (2019): ERA5-Land daily averaged data from 1981 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). (2023-03-30).

Messager, M.L., Lehner, B., Grill, G., Nedeva, I., Schmitt, O. (2016). Estimating the volume and age of water stored in global lakes using a geo-statistical approach. Nature Communications, 7: 13603. <https://doi.org/10.1038/ncomms13603>

Masek, J.G., Vermote, E.F., Saleous N.E., Wolfe, R., Hall, F.G., Huemmrich, K.F., Gao, F., Kutler, J., and Lim, T-K. (2006). A Landsat surface reflectance dataset for North America, 1990--2000. IEEE Geoscience and Remote Sensing Letters 3(1):68-72. <http://dx.doi.org/10.1109/LGRS.2005.857030>.

R Core Team (2023). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL <https://www.R-project.org/>.

Topp, S. N., Pavelsky, T. M., Dugan, H. A., Yang, X., Gardner, J., & Ross, M. R. V. (2021). Shifting patterns of summer lake color phenology in over 26,000 US lakes. *Water Resources Research*, 57, e2020WR029123. <https://doi.org/10.1029/2020WR029123>

Vermote, E., Justice, C., Claverie, M., & Franch, B. (2016). Preliminary analysis of the performance of the Landsat 8/OLI land surface reflectance product. Remote Sensing of Environment. <http://dx.doi.org/10.1016/j.rse.2016.04.008>.

Yang, Xiao. (2020). Deepest point calculation for any given polygon using Google Earth Engine JavaScript API (Version v2). Zenodo. <https://doi.org/10.5281/zenodo.6341960>
