---
title: "xgboost algorithm for Secchi - high data stringency, high secchi focus"
author: "B Steele w/Edits from Matt Ross"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, comment = FALSE, message = FALSE,
                      cache = FALSE)

library(tidyverse)
library(xgboost)
library(Metrics)
library(ggpmisc)
library(ggthemes)

match_dir = 'data/matchups/'
model_dir = 'data/models/'


# Set random seed
set.seed(799)
```

# Purpose

The purpose of this script is to apply the `xgboost` algorithm to Remote
Sensing Imagery of Lake Yojoa in Honduras, to estimate Yojoa water
clarity. You can read more about this lake
[here](https://www.sciencedirect.com/science/article/pii/S0048969722015479).

In this very stringent approach, we subset the Secchi-Landsat matchups
by Landsat flyover, resulting in no image-date crossover between the
train, test, or validation datasets. We then
create a weighted feature that awards the model for estimating higher
Secchi values well disproportionate to lower Secchi values.

## Load matchup data

We'll load the environment variables from script 9 here to compare apples-to-apples.

```{r}
load('data/models/train_test_val_verystringent.RData')
```

We want to predict the `secchi` value in these datasets, so let's set
the `target` as that variable:

```{r}
## Identify our target (value is secchi)
target <- 'secchi'
```

# xgboost Runs

Here, we develop models for 3 days, 5 days, and jemma-special (7 days
except 1 day in oct/nov) matchup datasets.

## Name feature groups

Here, we indicate the features to be used in our models. We'll use the
visual bands and add in summaries of ERA5 met data. In our datasets, the
5-day met summaries have the suffix '\_5', etc.

```{r}
band_met3_feats <-  c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_3', 'max_temp_degK_3', 'mean_temp_degK_3', 'min_temp_degK_3',
                     'tot_precip_m_3', 'mean_wind_mps_3')

band_met5_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_5', 'max_temp_degK_5', 'mean_temp_degK_5', 'min_temp_degK_5',
                     'tot_precip_m_5', 'mean_wind_mps_5')

band_met51_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_5', 'max_temp_degK_5', 'mean_temp_degK_5', 'min_temp_degK_5',
                     'tot_precip_m_5', 'mean_wind_mps_5',
                     'solar_rad_KJpm2_prev', 'precip_m_prev','air_temp_degK_prev','wind_speed_mps_prev')

band_met71_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_7', 'max_temp_degK_7', 'mean_temp_degK_7', 'min_temp_degK_7',
                     'tot_precip_m_7', 'mean_wind_mps_7',
                     'solar_rad_KJpm2_prev', 'precip_m_prev','air_temp_degK_prev','wind_speed_mps_prev')
```

## Format data for xgboost

### 3 Day Window

```{r}
## 3 day window, 3 days previous met
dtrain_3d_3m <- xgb.DMatrix(data = as.matrix(train_3[,band_met3_feats]), 
                            label = train_3[,target],
                            weight = train_3$weight)

dtest_3d_3m <- xgb.DMatrix(data = as.matrix(test_3[,band_met3_feats]), 
                     label = test_3[,target],
                            weight = test_3$weight)

dval_3d_3m <- xgb.DMatrix(data = as.matrix(val_3[,band_met3_feats]), 
                     label = val_3[,target],
                            weight = val_3$weight)

## 5 day window, 5 days previous met
dtrain_3d_5m <- xgb.DMatrix(data = as.matrix(train_3[,band_met5_feats]), 
                            label = train_3[,target],
                            weight = train_3$weight)

dtest_3d_5m <- xgb.DMatrix(data = as.matrix(test_3[,band_met5_feats]), 
                     label = test_3[,target],
                            weight = test_3$weight)

dval_3d_5m <- xgb.DMatrix(data = as.matrix(val_3[,band_met5_feats]), 
                     label = val_3[,target],
                            weight = val_3$weight)

## 3 day window, 5/1 days previous met
dtrain_3d_51m <- xgb.DMatrix(data = as.matrix(train_3[,band_met51_feats]), 
                            label = train_3[,target])

dtest_3d_51m <- xgb.DMatrix(data = as.matrix(test_3[,band_met51_feats]), 
                     label = test_3[,target])

dval_3d_51m <- xgb.DMatrix(data = as.matrix(val_3[,band_met51_feats]), 
                     label = val_3[,target])

## 5 day window, 7/1 days previous met
dtrain_3d_71m <- xgb.DMatrix(data = as.matrix(train_3[,band_met71_feats]), 
                            label = train_3[,target],
                            weight = train_3$weight)

dtest_3d_71m <- xgb.DMatrix(data = as.matrix(test_3[,band_met71_feats]), 
                     label = test_3[,target],
                            weight = test_3$weight)

dval_3d_71m <- xgb.DMatrix(data = as.matrix(val_3[,band_met71_feats]), 
                     label = val_3[,target],
                            weight = val_3$weight)

```

### 5 Day Window

```{r}
## 5 day window, 3 days previous met
dtrain_5d_3m <- xgb.DMatrix(data = as.matrix(train_5[,band_met3_feats]), 
                            label = train_5[,target],
                            weight = train_5$weight)

dtest_5d_3m <- xgb.DMatrix(data = as.matrix(test_5[,band_met3_feats]), 
                     label = test_5[,target],
                            weight = test_5$weight)

dval_5d_3m <- xgb.DMatrix(data = as.matrix(val_5[,band_met3_feats]), 
                     label = val_5[,target],
                            weight = val_5$weight)

## 5 day window, 5 days previous met
dtrain_5d_5m <- xgb.DMatrix(data = as.matrix(train_5[,band_met5_feats]), 
                            label = train_5[,target],
                            weight = train_5$weight)

dtest_5d_5m <- xgb.DMatrix(data = as.matrix(test_5[,band_met5_feats]), 
                     label = test_5[,target],
                            weight = test_5$weight)

dval_5d_5m <- xgb.DMatrix(data = as.matrix(val_5[,band_met5_feats]), 
                     label = val_5[,target],
                            weight = val_5$weight)

## 5 day window, 5/1 days previous met
dtrain_5d_51m <- xgb.DMatrix(data = as.matrix(train_5[,band_met51_feats]), 
                            label = train_5[,target])

dtest_5d_51m <- xgb.DMatrix(data = as.matrix(test_5[,band_met51_feats]), 
                     label = test_5[,target])

dval_5d_51m <- xgb.DMatrix(data = as.matrix(val_5[,band_met51_feats]), 
                     label = val_5[,target])

## 5 day window, 7/1 days previous met
dtrain_5d_71m <- xgb.DMatrix(data = as.matrix(train_5[,band_met71_feats]), 
                            label = train_5[,target],
                            weight = train_5$weight)

dtest_5d_71m <- xgb.DMatrix(data = as.matrix(test_5[,band_met71_feats]), 
                     label = test_5[,target],
                            weight = test_5$weight)

dval_5d_71m <- xgb.DMatrix(data = as.matrix(val_5[,band_met71_feats]), 
                     label = val_5[,target],
                            weight = val_5$weight)

```

### Local Knowledge Window

```{r}
## jemma window, 5 days previous met
dtrain_jd_5m <- xgb.DMatrix(data = as.matrix(train_j[,band_met5_feats]), 
                            label = train_j[,target],
                            weight = train_j$weight)

dtest_jd_5m <- xgb.DMatrix(data = as.matrix(test_j[,band_met5_feats]), 
                     label = test_j[,target],
                            weight = test_j$weight)

dval_jd_5m <- xgb.DMatrix(data = as.matrix(val_j[,band_met5_feats]), 
                     label = val_j[,target],
                            weight = val_j$weight)

## jemma window, 3 days previous met
dtrain_jd_3m <- xgb.DMatrix(data = as.matrix(train_j[,band_met3_feats]), 
                            label = train_j[,target],
                            weight = train_j$weight)

dtest_jd_3m <- xgb.DMatrix(data = as.matrix(test_j[,band_met3_feats]), 
                     label = test_j[,target],
                            weight = test_j$weight)

dval_jd_3m <- xgb.DMatrix(data = as.matrix(val_j[,band_met3_feats]), 
                     label = val_j[,target],
                            weight = val_j$weight)

## jemma window, 5/1 days previous met
dtrain_jd_51m <- xgb.DMatrix(data = as.matrix(train_j[,band_met51_feats]), 
                            label = train_j[,target])

dtest_jd_51m <- xgb.DMatrix(data = as.matrix(test_j[,band_met51_feats]), 
                     label = test_j[,target])

dval_jd_51m <- xgb.DMatrix(data = as.matrix(val_j[,band_met51_feats]), 
                     label = val_j[,target])

## 5 day window, 7/1 days previous met
dtrain_jd_71m <- xgb.DMatrix(data = as.matrix(train_j[,band_met71_feats]), 
                            label = train_j[,target],
                            weight = train_j$weight)

dtest_jd_71m <- xgb.DMatrix(data = as.matrix(test_j[,band_met71_feats]), 
                     label = test_j[,target],
                            weight = test_j$weight)

dval_jd_71m <- xgb.DMatrix(data = as.matrix(val_j[,band_met71_feats]), 
                     label = val_j[,target],
                            weight = val_j$weight)

```

### Parameter optimization

This is an xgboost optimization method developed by Sam Sillen where you
list many possible hyperparameter options and then create a matrix of
all possible combinations and grab the top 20 performing combinations of
hyperparameters by square error (our loss statistic).

```{r}
# Hypertune xgboost parameters and save as 'best_params' 
grid_train <- expand.grid(
  max_depth= c(3,6,8),
  subsample = c(.5,.8,1),
  colsample_bytree= c(.5,.8,1),
  eta = c(0.1, 0.3),
  min_child_weight= c(3,5,7)
)

hypertune_xgboost = function(train,test, grid){
  params <- list(booster = "gbtree", objective = 'reg:squarederror', 
                 eta=grid$eta ,max_depth=grid$max_depth, 
                 min_child_weight=grid$min_child_weight,
                 subsample=grid$subsample, 
                 colsample_bytree=grid$colsample_bytree)
  xgb.naive <- xgb.train(params = params, data = train, nrounds = 1000, 
                         watchlist = list(train = train, val = test), 
                         verbose = 0,
                         early_stopping_rounds = 20)
  summary <- grid %>% mutate(val_loss = xgb.naive$best_score, best_message = xgb.naive$best_msg,
                             mod = list(xgb.naive))
  
  return(summary) 
}
```

### 3 day window, 3 day met hypertuning

Note, evaluation is turned off for these chunks as to not overwrite
previous models and parameter tuning in next section.

```{r, eval = F}
## Hypertune xgboost 3 day window, 3 day met
xgboost_hypertune_3d_3m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_3d_3m,dtest_3d_3m,current)
  })

mod_summary_3d_3m <- xgboost_hypertune_3d_3m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_3d_3m <- xgboost_hypertune_3d_3m[xgboost_hypertune_3d_3m$val_loss==min(xgboost_hypertune_3d_3m$val_loss),]

save(mod_summary_3d_3m,best_mod_3d_3m, file = 'data/models/paramsxg_ep_ds_val_3d_3m.RData')
```

### 3 day window, 5 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 3 day window, 5 day met
xgboost_hypertune_3d_5m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_3d_5m,dtest_3d_5m,current)
  })

mod_summary_3d_5m <- xgboost_hypertune_3d_5m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_3d_5m <- xgboost_hypertune_3d_5m[xgboost_hypertune_3d_5m$val_loss==min(xgboost_hypertune_3d_5m$val_loss),]

save(mod_summary_3d_5m,best_mod_3d_5m, file = 'data/models/paramsxg_ep_ds_val_3d_5m.RData')

```

### 3 day window, 5/1 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 5/1 day met
xgboost_hypertune_3d_51m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_3d_51m,dtest_3d_51m,current)
  })

mod_summary_3d_51m <- xgboost_hypertune_3d_51m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_3d_51m <- xgboost_hypertune_3d_51m[xgboost_hypertune_3d_51m$val_loss==min(xgboost_hypertune_3d_51m$val_loss),]

save(mod_summary_3d_51m,best_mod_3d_51m, file = 'data/models/paramsxg_ep_ds_val_3d_51m.RData')
```

### 3 day window, 7/1 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 7/1 day met
xgboost_hypertune_3d_71m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_3d_71m,dtest_3d_71m,current)
  })

mod_summary_3d_71m <- xgboost_hypertune_3d_71m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_3d_71m <- xgboost_hypertune_3d_71m[xgboost_hypertune_3d_71m$val_loss==min(xgboost_hypertune_3d_71m$val_loss),]

save(mod_summary_3d_71m,best_mod_3d_71m, file = 'data/models/paramsxg_ep_ds_val_3d_71m.RData')
```

### 5 day window, 3 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 3 day met
xgboost_hypertune_5d_3m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_5d_3m,dtest_5d_3m,current)
  })

mod_summary_5d_3m <- xgboost_hypertune_5d_3m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_5d_3m <- xgboost_hypertune_5d_3m[xgboost_hypertune_5d_3m$val_loss==min(xgboost_hypertune_5d_3m$val_loss),]

save(mod_summary_5d_3m,best_mod_5d_3m, file = 'data/models/paramsxg_ep_ds_val_5d_3m.RData')
```

### 5 day window, 5 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 5 day met
xgboost_hypertune_5d_5m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_5d_5m,dtest_5d_5m,current)
  })

mod_summary_5d_5m <- xgboost_hypertune_5d_5m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_5d_5m <- xgboost_hypertune_5d_5m[xgboost_hypertune_5d_5m$val_loss==min(xgboost_hypertune_5d_5m$val_loss),]

save(mod_summary_5d_5m,best_mod_5d_5m, file = 'data/models/paramsxg_ep_ds_val_5d_5m.RData')

```

### 5 day window, 5/1 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 5/1 day met
xgboost_hypertune_5d_51m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_5d_51m,dtest_5d_51m,current)
  })

mod_summary_5d_51m <- xgboost_hypertune_5d_51m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_5d_51m <- xgboost_hypertune_5d_51m[xgboost_hypertune_5d_51m$val_loss==min(xgboost_hypertune_5d_51m$val_loss),]

save(mod_summary_5d_51m,best_mod_5d_51m, file = 'data/models/paramsxg_ep_ds_val_5d_51m.RData')
```

### 5 day window, 7/1 day met hypertuning

```{r, eval = F}
## Hypertune xgboost 5 day window, 7/1 day met
xgboost_hypertune_5d_71m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_5d_71m,dtest_5d_71m,current)
  })

mod_summary_5d_71m <- xgboost_hypertune_5d_71m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_5d_71m <- xgboost_hypertune_5d_71m[xgboost_hypertune_5d_71m$val_loss==min(xgboost_hypertune_5d_71m$val_loss),]

save(mod_summary_5d_71m,best_mod_5d_71m, file = 'data/models/paramsxg_ep_ds_val_5d_71m.RData')
```

### Jemma-special, 3 day met

```{r, eval = F}
## Hypertune xgboost jemma window, 3 day met
xgboost_hypertune_jd_3m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_jd_3m,dtest_jd_3m,current)
  })

mod_summary_jd_3m <- xgboost_hypertune_jd_3m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_jd_3m <- xgboost_hypertune_jd_3m[xgboost_hypertune_jd_3m$val_loss==min(xgboost_hypertune_jd_3m$val_loss),]

save(mod_summary_jd_3m,best_mod_jd_3m, file = 'data/models/paramsxg_ep_ds_val_jd_3m.RData')
```

### Jemma-special, 5 day met

```{r, eval = F}
## Hypertune xgboost jemma window, 5 day met
xgboost_hypertune_jd_5m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_jd_5m,dtest_jd_5m,current)
  })

mod_summary_jd_5m <- xgboost_hypertune_jd_5m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_jd_5m <- xgboost_hypertune_jd_5m[xgboost_hypertune_jd_5m$val_loss==min(xgboost_hypertune_jd_5m$val_loss),]

save(mod_summary_jd_5m,best_mod_jd_5m, file = 'data/models/paramsxg_ep_ds_val_jd_5m.RData')
```

### Jemma-special, 5/1 day met

```{r, eval = F}
## Hypertune xgboost jemma window, 5/1 day met
xgboost_hypertune_jd_51m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_jd_51m,dtest_jd_51m,current)
  })

mod_summary_jd_51m <- xgboost_hypertune_jd_51m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_jd_51m <- xgboost_hypertune_jd_51m[xgboost_hypertune_jd_51m$val_loss==min(xgboost_hypertune_jd_51m$val_loss),]

save(mod_summary_jd_51m,best_mod_jd_51m, file = 'data/models/paramsxg_ep_ds_val_jd_51m.RData')
```

### Jemma-special, 7/1 day met

```{r, eval = F}
## Hypertune xgboost jemma window, 5 day met
xgboost_hypertune_jd_71m <- grid_train %>%
  pmap_dfr(function(...) {
    current <- tibble(...)
    hypertune_xgboost(dtrain_jd_71m,dtest_jd_71m,current)
  })

mod_summary_jd_71m <- xgboost_hypertune_jd_71m %>% 
  arrange(val_loss) %>%
  dplyr::slice(1:20)

best_mod_jd_71m <- xgboost_hypertune_jd_71m[xgboost_hypertune_jd_71m$val_loss==min(xgboost_hypertune_jd_71m$val_loss),]

save(mod_summary_jd_71m,best_mod_jd_71m, file = 'data/models/paramsxg_ep_ds_val_jd_71m.RData')
```


# Model Assessment and Application

### load model summaries

```{r}
load('data/models/paramsxg_ep_ds_val_3d_5m.RData')
load('data/models/paramsxg_ep_ds_val_3d_3m.RData')
load('data/models/paramsxg_ep_ds_val_3d_51m.RData')
load('data/models/paramsxg_ep_ds_val_3d_71m.RData')
load('data/models/paramsxg_ep_ds_val_5d_5m.RData')
load('data/models/paramsxg_ep_ds_val_5d_3m.RData')
load('data/models/paramsxg_ep_ds_val_5d_51m.RData')
load('data/models/paramsxg_ep_ds_val_5d_71m.RData')
load('data/models/paramsxg_ep_ds_val_jd_5m.RData')
load('data/models/paramsxg_ep_ds_val_jd_3m.RData')
load('data/models/paramsxg_ep_ds_val_jd_51m.RData')
load('data/models/paramsxg_ep_ds_val_jd_71m.RData')

```

Now that these are loaded, we need to look at the test/train statistics.
Ideally the train/test RMSE are relatively close so we don't choose too
overfit of a model. Below, we apply the model to the validation dataset and plot the validation observed versus predicted.

The optimized booster was chosen as the highest-ranked booster where the train/test were within 0.15m RMSE. If no booster met that, the booster with the closest train/test RMSE was chosen.

### Three day window dataset

```{r}
mod_summary_3d_3m %>% select(best_message)
mod_summary_3d_5m %>% select(best_message)
mod_summary_3d_51m %>% select(best_message)
mod_summary_3d_71m %>% select(best_message)
```

```{r}
# most of best models are overfit, so looking for train/test RMSE that are closer
optimized_booster_3d_3m <- mod_summary_3d_3m$mod[1][[1]]
optimized_booster_3d_5m <- mod_summary_3d_5m$mod[1][[1]]
optimized_booster_3d_51m <- mod_summary_3d_51m$mod[8][[1]]
optimized_booster_3d_71m <- mod_summary_3d_71m$mod[15][[1]]
#note that pretty much all of these are overfit.

# Apply best mod
preds_3 <- val_3 %>% 
  mutate(pred_secchi_3d_5m = predict(optimized_booster_3d_5m, dval_3d_5m),
         pred_secchi_3d_3m = predict(optimized_booster_3d_3m, dval_3d_3m),
         pred_secchi_3d_51m = predict(optimized_booster_3d_51m, dval_3d_51m),
         pred_secchi_3d_71m = predict(optimized_booster_3d_71m, dval_3d_71m))

evals_3 <- preds_3 %>%
  summarise(across(c(pred_secchi_3d_3m, pred_secchi_3d_5m, pred_secchi_3d_71m, pred_secchi_3d_51m),
                   list(rmse = ~rmse(secchi, .),
                        mae = ~mae(secchi, .),
                        mape = ~mape(secchi, .),
                        bias = ~bias(secchi, .),
                        p.bias = ~percent_bias(secchi, .),
                        smape = ~smape(secchi, .),
                        r2 = ~cor(secchi, .)^2), 
                   .names = "{fn}_{col}"))

evals_3
```

### Five day window dataset

```{r}
mod_summary_5d_3m %>% select(best_message)
mod_summary_5d_5m %>% select(best_message)
mod_summary_5d_51m %>% select(best_message)
mod_summary_5d_71m %>% select(best_message)
```

```{r}
# most of best models are overfit, so looking for train/test RMSE that are closer
optimized_booster_5d_3m <- mod_summary_5d_3m$mod[14][[1]]
optimized_booster_5d_5m <- mod_summary_5d_5m$mod[16][[1]]
optimized_booster_5d_51m <- mod_summary_5d_51m$mod[8][[1]]
optimized_booster_5d_71m <- mod_summary_5d_71m$mod[3][[1]]

# Apply best mod
preds_5 <- val_5 %>% 
  mutate(pred_secchi_5d_5m = predict(optimized_booster_5d_5m, dval_5d_5m),
         pred_secchi_5d_3m = predict(optimized_booster_5d_3m, dval_5d_3m),
         pred_secchi_5d_51m = predict(optimized_booster_5d_51m, dval_5d_51m),
         pred_secchi_5d_71m = predict(optimized_booster_5d_71m, dval_5d_71m))

evals_5 <- preds_5 %>%
  summarise(across(c(pred_secchi_5d_3m, pred_secchi_5d_5m, pred_secchi_5d_71m,pred_secchi_5d_51m),
                   list(rmse = ~rmse(secchi, .),
                        mae = ~mae(secchi, .),
                        mape = ~mape(secchi, .),
                        bias = ~bias(secchi, .),
                        p.bias = ~percent_bias(secchi, .),
                        smape = ~smape(secchi, .),
                        r2 = ~cor(secchi, .)^2), 
                   .names = "{fn}_{col}"))

evals_5
```

### Jemma-special window dataset

```{r}
mod_summary_jd_3m %>% select(best_message)
mod_summary_jd_5m %>% select(best_message)
mod_summary_jd_51m %>% select(best_message)
mod_summary_jd_71m %>% select(best_message)
```

```{r}
optimized_booster_jd_3m <- mod_summary_jd_3m$mod[17][[1]]
optimized_booster_jd_5m <- mod_summary_jd_5m$mod[18][[1]]
optimized_booster_jd_51m <- mod_summary_jd_51m$mod[3][[1]]
optimized_booster_jd_71m <- mod_summary_jd_71m$mod[18][[1]]

# Apply best mod
preds_jd <- val_j %>% 
  mutate(pred_secchi_jd_5m = predict(optimized_booster_jd_5m, dval_jd_5m),
         pred_secchi_jd_3m = predict(optimized_booster_jd_3m, dval_jd_3m),
         pred_secchi_jd_51m = predict(optimized_booster_jd_51m, dval_jd_51m),
         pred_secchi_jd_71m = predict(optimized_booster_jd_71m, dval_jd_71m))

evals_jd <- preds_jd %>%
  summarise(across(c(pred_secchi_jd_3m, pred_secchi_jd_5m, pred_secchi_jd_71m, pred_secchi_jd_51m),
                   list(rmse = ~rmse(secchi, .),
                        mae = ~mae(secchi, .),
                        mape = ~mape(secchi, .),
                        bias = ~bias(secchi, .),
                        p.bias = ~percent_bias(secchi, .),
                        smape = ~smape(secchi, .),
                        r2 = ~cor(secchi, .)^2), 
                   .names = "{fn}_{col}"))

evals_jd
```

### Model Performance - 3 day window

Keep in mind that all of these models seemed overfit in the train/test sets.

```{r, echo = F}
ggplot(preds_3, aes(x = secchi, y = pred_secchi_3d_3m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nthree day matchups, band and 3-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_3, aes(x = secchi, y = pred_secchi_3d_5m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nthree day matchups, band and 5-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_3, aes(x = secchi, y = pred_secchi_3d_51m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nthree day matchups, band and 5/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_3, aes(x = secchi, y = pred_secchi_3d_71m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nthree day matchups, band and 7/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))
```

### Model Performance - 5 day window

```{r}
ggplot(preds_5, aes(x = secchi, y = pred_secchi_5d_3m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nfive day matchups, band and 3-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_5, aes(x = secchi, y = pred_secchi_5d_5m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nfive day matchups, band and 5-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_5, aes(x = secchi, y = pred_secchi_5d_51m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nfive day matchups, band and 5/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_5, aes(x = secchi, y = pred_secchi_5d_71m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\nfive day matchups, band and 7/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))
```

### Model Performance - Jemma window

```{r}
ggplot(preds_jd, aes(x = secchi, y = pred_secchi_jd_3m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\n7/1 day matchups, band and 3-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_jd, aes(x = secchi, y = pred_secchi_jd_5m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\n7/1 day matchups, band and 5-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_jd, aes(x = secchi, y = pred_secchi_jd_51m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\n7/1 day matchups, band and 5/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

ggplot(preds_jd, aes(x = secchi, y = pred_secchi_jd_71m)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Yojoa Secchi xgboost Validation Data\n7/1 day matchups, band and 7/1-day met summaries', 
       subtitle = 'high data stringency\nweighted Secchi\ngrey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))
```

These all look terrible.

# Applying model to full data

Local knowledge window does not perform well in this scenario. Best performing seems to be 5-day matchup and 3-day met summary.

```{r}
features = band_met3_feats
model = optimized_booster_5d_3m
met = '3 day met summaries'
window = '5 day window'

full_stack <- read_csv('data/upstreamRS/yojoa_corr_rrs_met_scaled_v2023-06-15.csv') %>%
  mutate(secchi = 100) %>% 
    mutate(secchi = as.numeric(secchi)) %>% #there's one wonky value in here with two decimal points... dropping from this analysis
    filter(!is.na(secchi))

stack_xgb <- xgb.DMatrix(data = as.matrix(full_stack[,features]))

full_stack_simp <- full_stack %>%
  mutate(secchi = predict(model, stack_xgb)) %>%
  select(date, location, secchi, mission) 

situ_stack <- read_csv('data/in-situ/Secchi_completedataset.csv') %>%
  mutate(secchi = as.numeric(secchi),
         date = mdy(date)) %>%
  filter(!is.na(secchi)) %>%
  mutate(mission = 'Measured') %>%
  bind_rows(full_stack_simp)%>% 
  mutate(location = gsub(' ', '', location))
```

Let's look at each of the site records alongside the Landsat-estimated Secchi depth.

```{r, echo = F}
plotRecordBySite = function(site) {
  ggplot(situ_stack %>%
           filter(location == site), aes(x = date, y = secchi, color = mission,
                                        shape = mission)) + 
    geom_point() + 
    labs(title = paste0('Yojoa Secchi Historical Record - Site ', site),
         subtitle = paste0(window, ', ', met, '\nhigh data stringency, weighted Secchi'),
         y = 'Secchi (m)',
         color = 'data source', shape = 'data source') +
    scale_color_manual(values = c('grey10','grey30','grey50','grey70','blue')) + 
    theme_few() +
    theme(legend.position = c(0.8,0.8)) + 
    scale_shape_manual(values = c(19,19,19,19,1)) +
    scale_y_continuous(limits = c(0, max(situ_stack$secchi)), breaks = seq(0, max(situ_stack$secchi), 2)) +
    scale_x_date(limits = c(min(situ_stack$date), max(situ_stack$date))) +
    theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
          plot.subtitle = element_text(hjust = 0.5))
}

map(sort(unique(situ_stack$location)), plotRecordBySite)
```

### Look at recent data per location

```{r}
plotRecentBySite = function(site) {
  ggplot(situ_stack %>%
           filter(location == site), aes(x = date, y = secchi, color = mission,
                                        shape = mission)) + 
    geom_point() + 
    labs(title = paste0('Yojoa Secchi 2018-2022 - Site ', site),
         subtitle = paste0(window, ', ', met, '\nhigh data stringency, weighted Secchi'),
         y = 'Secchi (m)',
         color = 'data source', shape = 'data source') +
    scale_color_manual(values = c('grey10','grey30','grey50','grey70','blue')) + 
    theme_few() +
    theme(legend.position = c(0.8,0.8)) + 
    scale_shape_manual(values = c(19,19,19,19,1)) +
    scale_y_continuous(limits = c(0, max(situ_stack$secchi)), breaks = seq(0, max(situ_stack$secchi), 2)) +
    scale_x_date(limits = c(ymd('2018-01-01'), max(situ_stack$date))) +
    theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
          plot.subtitle = element_text(hjust = 0.5))
}

map(sort(unique(situ_stack$location)), plotRecentBySite)
```


## Whole lake secchi dynamics

While there is plenty of variability across the lake, let's summarize to a single value per date, since not all sites have the same density of record. Since there are a few oddballs in here (both in terms of measured and estimated), we'll use the median Secchi across all sites.

```{r}
lake_med <- situ_stack %>%
  group_by(date,mission) %>%
  summarize(across(where(is.numeric),median))
```

```{r, echo = F}
ggplot(lake_med, aes(x = date, y = secchi, color = mission, shape = mission)) + 
  geom_point() + 
  scale_color_manual(values = c('grey10','grey30','grey50','grey70','blue')) + 
  labs(title = 'Yojoa Secchi Historical Record\nwhole-lake median',
         subtitle = paste0(window, ', ', met, '\nhigh data stringency, weighted Secchi'),
       y = 'median Secchi(m)',
       color = 'data source', shape = 'data source') +
  theme_few() +
  theme(legend.position = c(0.8,0.8)) + 
  scale_shape_manual(values = c(19,19,19,19,1)) +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

```

### 2006

```{r, echo = F}
ggplot(lake_med, aes(x = date, y = secchi, color = mission,shape = mission)) + 
  geom_point() + 
  scale_color_manual(values = c('grey10','grey30','grey50','grey70','blue')) + 
  theme_few() +
  labs(title = 'Yojoa Secchi 2006\nwhole-lake median',
         subtitle = paste0(window, ', ', met, '\nhigh data stringency, weighted Secchi'),
       y = 'median Secchi (m)',
       color = 'data source', shape = 'data source') +
  theme(legend.position = c(0.8,0.8)) + 
  scale_shape_manual(values = c(19,19,19,19,1)) +
  scale_x_date(limits = c(as.Date('2006-01-01'), as.Date('2006-12-31')))+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

```

### Recent

```{r}
ggplot(lake_med, aes(x = date, y = secchi, color = mission, shape = mission)) + 
  geom_point() + 
  scale_color_manual(values = c('grey10','grey30','grey50','grey70','blue')) + 
  theme_few() +
  labs(title = 'Yojoa Secchi 2018-2022\nwhole-lake median',
         subtitle = paste0(window, ', ', met, '\nhigh data stringency, weighted Secchi'),
       y = 'median Secchi (m)',
       color = 'data source', shape = 'data source') +
  theme(legend.position = c(0.8,0.8)) + 
  scale_shape_manual(values = c(19,19,19,19,1)) +
  scale_x_date(limits = c(as.Date('2018-01-01'), as.Date('2023-01-01')))+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))

```

## Print off xgboost model feature importance

```{r}
feat_importance = xgb.importance(feature_names = band_met51_feats,
               model = optimized_booster_jd_51m)

xgb.plot.importance(feat_importance)
```
