---
title: "Calculate Estimated Secchi"
author: "B Steele, Matt Ross"
date: "2023-06-22"
output: html_document
---

```{r}
library(tidyverse)
library(caret)
library(xgboost)
library(lubridate)
library(readr)

match_dir = 'data/matchups/'
rs_dir = 'data/upstreamRS/'
is_dir = 'data/in-situ/'
est_dir = 'data/landsat_estimations/'
```

# Purpose

This script applies all of the top-performing models to the Landsat stack and exports the Landsat timerseries as a .csv file.

## Load data

```{r}
stack = read.csv(file.path(rs_dir, 'yojoa_corr_rrs_met_scaled_v2023-06-15.csv')) %>% 
  mutate(secchi = NA_real_) %>% 
  filter(date < ymd('2023-01-01')) #this is when met data ends
```

## Load models and parameters

Here, we load the models. 

```{r}
load('data/models/optimized_xg_8_jd_51m.RData')
xg_8_71d_51m = optimized_booster_jd_51m
load('data/models/optimized_xg_9_5d_71m.RData')
xg_9_5d_71m = optimized_booster_5d_71m
load('data/models/optimized_xg_10_5d_3m.RData')
xg_10_5d_3m = optimized_booster_5d_3m
load('data/models/stepreg_71d_5m.RData')
```

And save the parameter groups

```{r}
band_met3_feats <-  c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_3', 'max_temp_degK_3', 'mean_temp_degK_3', 'min_temp_degK_3',
                     'tot_precip_m_3', 'mean_wind_mps_3')

band_met51_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_5', 'max_temp_degK_5', 'mean_temp_degK_5', 'min_temp_degK_5',
                     'tot_precip_m_5', 'mean_wind_mps_5',
                     'solar_rad_KJpm2_prev', 'precip_m_prev','air_temp_degK_prev','wind_speed_mps_prev')

band_met71_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_7', 'max_temp_degK_7', 'mean_temp_degK_7', 'min_temp_degK_7',
                     'tot_precip_m_7', 'mean_wind_mps_7',
                     'solar_rad_KJpm2_prev', 'precip_m_prev','air_temp_degK_prev','wind_speed_mps_prev')

band_met5_feats_scaled = c('med_Blue_corr_scaled', 'med_Red_corr_scaled', 'med_Green_corr_scaled', 
         'med_Nir_corr_scaled',
         'RN_scaled', 'BG_scaled', 'RB_scaled', 'GB_scaled', 
         "tot_sol_rad_KJpm2_5_scaled", "max_temp_degK_5_scaled", "mean_temp_degK_5_scaled", "min_temp_degK_5_scaled",
         "tot_precip_m_5_scaled", "mean_wind_mps_5_scaled")

```

## Apply to the dataset

### stringent {xgboost}

```{r}
model = xg_8_71d_51m
features = band_met51_feats

stack_8_xgb <- xgb.DMatrix(data = as.matrix(stack[,features]))

stack_8 <- stack %>%
  mutate(est_secchi_stringent = predict(model, stack_8_xgb)) %>%
  select(date, location, est_secchi_stringent, mission) 
```

### very stringent {xgboost}

```{r}
model = xg_9_5d_71m
features = band_met71_feats

stack_9_xgb <- xgb.DMatrix(data = as.matrix(stack[,features]))

stack_9 <- stack %>%
  mutate(est_secchi_verystringent = predict(model, stack_9_xgb)) %>%
  select(date, location, est_secchi_verystringent, mission)
```

### very stringent {xgboost} with weighted high Secchi

```{r}
model = xg_10_5d_3m
features = band_met3_feats

stack_10_xgb <- xgb.DMatrix(data = as.matrix(stack[,features]))

stack_10 <- stack %>%
  mutate(est_secchi_highsecchi = predict(model, stack_10_xgb)) %>%
  select(date, location, est_secchi_highsecchi, mission)
```

### stepwise regression

```{r}
stack_11 <- stack %>% 
  mutate(est_secchi_stepwise = predict.train(object = step.model.71d5m, newdata = stack)) %>% 
  select(date, location, est_secchi_stepwise, mission)
```

## save files

```{r}
all_est = full_join(stack_8, stack_9) %>% 
  full_join(., stack_10) %>% 
  full_join(., stack_11) %>% 
  relocate(date, location, mission) 
write_csv(all_est, file.path(est_dir, 'Yojoa_LS-derived_Secchi_estimates.csv'))
```

