---
title: "2_LSC2_secchi_matchup"
author: "B Steele"
date: "2023-05-26"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)

is_dir = 'data/in-situ/'
coll_dir = 'data/upstreamRS/'
match_dir = 'data/matchups/'

```

# Purpose

To create a match up database of the Landsat record and historical Secchi data for Yojoa.

# Bring in the data

```{r}
secchi = read.csv(file.path(is_dir, 'Secchi_completedataset.csv')) %>% 
  mutate(date = as.Date(date, format = '%m/%d/%y')) 
str(secchi)
#there are duplicated records in here, remove them
secchi = secchi[!duplicated(secchi),]

ls = read.csv(file.path(coll_dir, 'yojoa_corr_rrs_met_v2023-06-15.csv')) %>% 
  mutate(date = as.Date(date)) 
str(ls)
```

## Normalize the SR and met data

While unnecessary for the {xgboost} "gbtree" booster, we will scale the data for comparison with the stepwise regression method.

```{r}
ls = ls %>% 
  mutate(across(c(med_Blue_corr:med_Swir2_corr, tot_sol_rad_KJpm2_3:wind_speed_mps_prev),
                scale, 
                .names = "{col}_scaled")) 
names(ls)

```

# Create matchups

And join them together (this is the list of 1:1 matchups)

```{r}
sameDay_matchups = inner_join(secchi, ls)
```

And now we can move out from there:

```{r}
#force join by location (of all possible pairs by location)
over_match = secchi %>% 
  rename(obs_date = date) %>% 
  full_join(ls, secchi, by = 'location') %>% 
  mutate(date_diff = obs_date-date) %>%  #calculate the time difference
  relocate(obs_date, date, location, date_diff)

# in order to not overmatch on either the in-situ location or the satellite (we want distinct, closest matches for each of these) we need to iteratively summarize to the closest match per satellite per observation data and location, then limit only one satelite image per location and flyover date
oneDay_matchups = over_match %>% 
  filter(abs(date_diff) <= days(1))%>% 
  arrange(abs(date_diff)) %>% 
  group_by(obs_date, location, mission) %>% 
  summarize_all(first)%>% 
  group_by(date, location) %>% 
  summarize_all(first)
twoDay_matchups = over_match %>% 
  filter(abs(date_diff) <= days(2))%>% 
  arrange(abs(date_diff)) %>% 
  group_by(obs_date, location, mission) %>% 
  summarize_all(first)%>% 
  group_by(date, location) %>% 
  summarize_all(first)
threeDay_matchups = over_match %>% 
  filter(abs(date_diff) <= days(3))%>% 
  arrange(abs(date_diff)) %>% 
  group_by(obs_date, location, mission) %>% 
  summarize_all(first)%>% 
  group_by(date, location) %>% 
  summarize_all(first)
fourDay_matchups = over_match %>% 
  filter(abs(date_diff) <= days(4))%>% 
  arrange(abs(date_diff)) %>% 
  group_by(obs_date, location, mission) %>% 
  summarize_all(first)%>% 
  group_by(date, location) %>% 
  summarize_all(first)
fiveDay_matchups = over_match %>% 
  filter(abs(date_diff) <= days(5))%>% 
  arrange(abs(date_diff)) %>% 
  group_by(obs_date, location, mission) %>% 
  summarize_all(first) %>% 
  group_by(date, location) %>% 
  summarize_all(first)

#per local knowledge, oct and nov can have very immediate changes, and things are pretty stable otherwise.
multi_match = over_match %>% 
  mutate(month = format(obs_date, '%m')) %>% 
  filter((month %in% c(10,11) & abs(date_diff) < days(1) |
            !(month %in% c(10,11)) & abs(date_diff) < days(7))) %>% 
  arrange(abs(date_diff)) %>% 
  group_by(date, location) %>% 
  summarize_all(first) %>% 
  group_by(date, location) %>% 
  summarize_all(first)
```

## Export matchups

```{r}
write.csv(sameDay_matchups, file.path(match_dir, paste0('sameDay_LS-Secchi_matchups_n', nrow(sameDay_matchups), '.csv')), row.names = F)
write.csv(oneDay_matchups, file.path(match_dir, paste0('oneDay_LS-Secchi_matchups_n', nrow(oneDay_matchups), '.csv')), row.names = F)
write.csv(twoDay_matchups, file.path(match_dir, paste0('twoDay_LS-Secchi_matchups_n', nrow(twoDay_matchups), '.csv')), row.names = F)
write.csv(threeDay_matchups, file.path(match_dir, paste0('threeDay_LS-Secchi_matchups_n', nrow(threeDay_matchups), '.csv')), row.names = F)
write.csv(fiveDay_matchups, file.path(match_dir, paste0('fiveDay_LS-Secchi_matchups_n', nrow(fiveDay_matchups), '.csv')), row.names = F)
write.csv(multi_match, file.path(match_dir, paste0('jemma_multiDay_LS-Secchi_matchups_n', nrow(multi_match), '.csv')), row.names = F)
```
