---
title: "xgboost algorithm for Secchi"
author: "B Steele w/Edits from Matt Ross"
date: "2023-05-25"
output: html_document
---

```{r}
library(tidyverse)
library(xgboost)
library(Metrics)
library(ggpmisc)

match_dir = 'data/matchups/'
model_dir = 'data/models/'

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, comment = FALSE, message = FALSE,
                      cache = TRUE)

```

# Purpose

The original purpose of this script is to apply applying the `xgboost` algorithm 
to Remote Sensing Imagery of Lake Yojoa in Honduras, to estimate Yojoa water clarity. 
You can read more about this lake [here](https://www.sciencedirect.com/science/article/pii/S0048969722015479). We
have slightly adopted the code to become a teaching demo on how to use machine learning
algorithms. We also use a myriad of climate covariates 
from the ERA5 climate data in this analysis.

You can read more about xgboost all over the internet, but I like the 
kaggle [demo](https://www.kaggle.com/code/rtatman/machine-learning-with-xgboost-in-r/notebook)

## Load matchup data

```{r}
#list all the files in the match directory
match = list.files(match_dir)

#load the three day matchup file. 
sameDay = read.csv(file.path(match_dir, match[grepl('same', match)]))
oneDay = read.csv(file.path(match_dir, match[grepl('one', match)]))
threeDay = read.csv(file.path(match_dir, match[grepl('three', match)]))
fiveDay = read.csv(file.path(match_dir, match[grepl('five', match)]))


```

Prep the data for xgboost

```{r}
prepData = function(df) {
  #make a rowid column
  df_prep = df %>% 
    rowid_to_column() %>% 
    mutate(secchi = as.numeric(secchi)) %>% #there's one wonky value in here with two decimal points... dropping from this analysis
    filter(!is.na(secchi))
  
  #Add ratios then trim to needd to columns to speed up run
  df_prep %>% 
    mutate(RN= med_Red_corr/med_Nir_corr,
           BG= med_Blue_corr/med_Green_corr,
           RB= med_Red_corr/med_Blue_corr,
           GB = med_Green_corr/med_Blue_corr)
}


sameDay <- prepData(sameDay)
oneDay <- prepData(oneDay)
threeDay <- prepData(threeDay)
fiveDay <- prepData(fiveDay)
```

We want to predict the `secchi` value in these datasets, so let's set the `target` as that variable:

```{r}
## Identify our target (value is secchi)
target <- 'secchi'
```



## Quick xgboost run on fiveDay matchups

Let's see what happens if we loosen our time restraint and add more matchups into the mix

### Make test and training sets

For the same day matchup dataset, let's grab 20% of the data as the 'test' set and the remainder as the training set.

```{r}
# Set random seed
set.seed(799)

##Pull 20% as holdout test data
test_fiveDay <- fiveDay %>%
  sample_frac(.2) 

## Remove holdout data
train_fiveDay <- fiveDay %>% filter(!rowid %in% test_fiveDay$rowid) 

hist(train_fiveDay$secchi)
hist(test_fiveDay$secchi)
```


## Add in the met data with the five day matchups

Let's see what happens if we add in the ERA5 met data. For this example, we'll use the 5-day summaries, meaning we've summarized the met data as the mean of the previous 5 days. Since we already made the training/test datasets, let's stick with those, but name new features.

### xgboost on band data and all the 5-day met data

In our dataset, the 5-day met summaries have the suffix '\_5'

```{r}
band_met5_feats <- c('med_Blue_corr', 'med_Green_corr', 'med_Red_corr', 'med_Nir_corr',
                     'RN', 'BG', 'RB','GB',
                     'tot_sol_rad_KJpm2_5', 'max_temp_degK_5', 'min_temp_degK_5',
                     'tot_precip_m_5', 'mean_wind_mps_5')
```

Now we'll format the data

```{r}
## Format it the way xgboost likes
dtrain_fd_bm5 <- xgb.DMatrix(data = as.matrix(train_fiveDay[,band_met5_feats]), 
                      label = train_fiveDay[,target])
dtest_fd_bm5 <- xgb.DMatrix(data = as.matrix(test_fiveDay[,band_met5_feats]), 
                     label = test_fiveDay[,target])

```

And now train the model

```{r}
#set the parameters of the boost algo
params <- list(booster = "gbtree", 
               objective = "reg:squarederror", 
               eta=0.01, 
               gamma=0.5, 
               max_depth=3, 
               min_child_weight=2, 
               subsample=1, 
               colsample_bytree=1)

#run the boost algo with those settings
xgb_naive_fd_bm5 <- xgb.train(params = params, 
                             data = dtrain_fd_bm5, 
                             nrounds = 1000, 
                             watchlist = list(train = dtrain_fd_bm5, 
                                              val = dtest_fd_bm5), 
                             print_every_n = 25, 
                             early_stopping_rounds = 10, 
                             maximize = F)
```

Okay, the train and test are still pretty far apart (which means we're overfitted), but let's look at the data anyway

```{r}

preds_fd_bm5 <- test_fiveDay %>% mutate(predicted_fd_bm5 = predict(xgb_naive_fd_bm5, dtest_fd_bm5))

evals_fd_bm5 <- preds_fd_bm5 %>%
  summarise(rmse = rmse(secchi, predicted_fd_bm5),
            mae = mae(secchi, predicted_fd_bm5),
            mape = mape(secchi, predicted_fd_bm5),
            bias = bias(secchi, predicted_fd_bm5),
            p.bias = percent_bias(secchi, predicted_fd_bm5),
            smape = smape(secchi, predicted_fd_bm5)) 

```

And let's visualize the predictions:

```{r}
ggplot(preds_fd_bm5, aes(x = secchi, y = predicted_fd_bm5)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Quick xgboost - Yojoa Secchi\nfive day matchups, band and 5-day met summaries', 
       subtitle = 'Grey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))
```

This generally looks better than before (our adjusted r-squared is higher that without the met data) AND it's better than the one-day matchups. Let's play around with the hyper parameters and see what happens...

### Try some different hyperparmeters

```{r}
#set the parameters of the boost algo
params <- list(booster = "gbtree", 
               objective = "reg:squarederror", 
               eta=0.3, 
               gamma=2, 
               max_depth=5, 
               min_child_weight=4, 
               subsample=1, 
               colsample_bytree=1)

#run the boost algo with those settings
xgb_naive_fd_bm5_play <- xgb.train(params = params, 
                             data = dtrain_fd_bm5, 
                             nrounds = 1000, 
                             watchlist = list(train = dtrain_fd_bm5, 
                                              val = dtest_fd_bm5), 
                             print_every_n = 25, 
                             early_stopping_rounds = 10, 
                             maximize = F)


preds_fd_bm5_play <- test_fiveDay %>% 
  mutate(predicted_fd_bm5 = predict(xgb_naive_fd_bm5_play, dtest_fd_bm5))

evals_fd_bm5_play <- preds_fd_bm5_play %>%
  summarise(rmse = rmse(secchi, predicted_fd_bm5),
            mae = mae(secchi, predicted_fd_bm5),
            mape = mape(secchi, predicted_fd_bm5),
            bias = bias(secchi, predicted_fd_bm5),
            p.bias = percent_bias(secchi, predicted_fd_bm5),
            smape = smape(secchi, predicted_fd_bm5),
            r2 = cor(secchi, predicted_fd_bm5)^2) 


evals_fd_bm5_play

```




And let's visualize the predictions:

```{r}
ggplot(preds_fd_bm5_play, aes(x = secchi, y = predicted_fd_bm5)) + 
  geom_point() +
  geom_abline(color = 'grey', lty = 2) + 
  coord_cartesian(xlim = c(0, 6.5),
                  ylim = c(0,6.5)) +
  stat_poly_eq(aes(label = paste(after_stat(adj.rr.label))),
               formula = y~x, 
               parse = TRUE, 
               label.y = Inf, 
               vjust = 1.3) +
  labs(title = 'Quick xgboost - Yojoa Secchi\nfive day matchups, band and 5-day met summaries', 
       subtitle = 'Grey dashed line is 1:1', 
       x = 'Actual Secchi (m)', 
       y = 'Predicted Secchi (m)')  +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.subtitle = element_text(hjust = 0.5))
```
